---
title: "650 Project"
author: "Jess"
date: "11/28/2021"
output: html_document
---

```{r}
#load packages
#install.packages('ggplot2','dplyr','tidyverse','readxl','patchwork','lme4','effects',"devtools","emmeans","gridExtra","lmtest") #install packages

library.list<- c('ggplot2','readxl','dplyr','tidyverse','patchwork','lme4','effects','sjPlot',"emmeans","gridExtra") #load packages
sapply(library.list, library, character.only=T, quietly=T)
library(NHANES)
#----
dat<-NHANES
head(dat)
quantile(dat$BMI, na.rm = TRUE)
dat$bmi_quan<-ifelse(dat$BMI<21.58, 1,
                            ifelse(21.58<=dat$BMI & dat$BMI<25.98,2,
                                   ifelse(25.98<=dat$BMI & dat$BMI<30.89, 3,
                                   ifelse(dat$BMI>=30.89,4,NA))))
dat2 <- dat[,c("ID","BPSysAve","BMI","bmi_quan","Age","Gender","Race1","Education","DirectChol","Diabetes","PhysActiveDays","SmokeNow","TotChol","AlcoholYear")]

#target_pop: nonmissing outsome and POI
target_pop <- dat2[!is.na(dat2$BPSysAve)&!is.na(dat2$BMI),] #8487
#Complete cases for all variables
comp_cases <- target_pop[complete.cases(target_pop),]
```

```{r}
#install.packages("qwraps2")
library("qwraps2")
orig_opt <- options()$qwraps2_markup
options(qwraps2_markup = "markdown")
summarytab<-
  list("Average Systolic Blood Pressure" =
       list("mean (sd)" = ~ qwraps2::mean_sd(BPSysAve,na_rm=TRUE)),
       "BMI" =
       list("mean (sd)" = ~ qwraps2::mean_sd(BMI,na_rm=TRUE)),
       "Age" =
       list("min"       = ~ min(Age),
            "max"       = ~ max(Age),
            "mean (sd)" = ~ qwraps2::mean_sd(Age,na_rm=TRUE)),
       "Gender" =
       list("Male" = ~ qwraps2::n_perc0(Gender == "male",na_rm=TRUE),
            "Female"  = ~ qwraps2::n_perc0(Gender == "female",na_rm=TRUE)),
       "Race" =
       list("White" = ~ qwraps2::n_perc0(Race1 == "White",na_rm=TRUE),
            "Black"  = ~ qwraps2::n_perc0(Race1 == "Black",na_rm=TRUE),
            "Mexican"= ~ qwraps2::n_perc0(Race1 == "Mexican",na_rm=TRUE),
            "Hispanic"= ~ qwraps2::n_perc0(Race1 == "Hispanic",na_rm=TRUE),
            "Other" = ~ qwraps2::n_perc0(Race1 == "Other",na_rm=TRUE)),
        "Eduaction" =
       list("8th Grade" = ~ qwraps2::n_perc0(Education == "8th Grade",na_rm=TRUE),
            "9 - 11th Grade"  = ~ qwraps2::n_perc0(Education == "9 - 11th Grade",na_rm=TRUE),
            "High School"= ~ qwraps2::n_perc0(Education == "High School",na_rm=TRUE),
            "Some College"= ~ qwraps2::n_perc0(Education == "Some College",na_rm=TRUE),
            "College Grad" = ~ qwraps2::n_perc0(Education == "College Grad",na_rm=TRUE),
            "Missing" = ~ qwraps2::n_perc0(is.na(Education))),
       "HDL Cholesterol" =
       list("mean (sd)" = ~ qwraps2::mean_sd(DirectChol,na_rm=TRUE)),
       "Diabetes" =
       list("Has diabetes" = ~ qwraps2::n_perc0(Diabetes == "Yes",na_rm=TRUE),
            "Doesn't have diabetes"  = ~ qwraps2::n_perc0(Diabetes == "No",na_rm=TRUE),
            "Missing" = ~ qwraps2::n_perc0(is.na(Diabetes))),
       "Physical Activity Days " =
       list("mean (sd)" = ~ qwraps2::mean_sd(PhysActiveDays,na_rm=TRUE)),
        "Smoking" =
       list("Smoker" = ~ qwraps2::n_perc0(SmokeNow == "Yes",na_rm=TRUE),
            "Nonsmoker"  = ~ qwraps2::n_perc0(SmokeNow == "No",na_rm=TRUE),
            "Missing" = ~ qwraps2::n_perc0(is.na(SmokeNow))),
        "Alcohol use per year"=
            list("mean (sd)" = ~ qwraps2::mean_sd(AlcoholYear,na_rm=TRUE))
       )
#only for those with Systolic Blood Pressure 
overall<-summary_table(comp_cases,summarytab)
bmi_quan<-summary_table(comp_cases,summarytab, by=c("bmi_quan"))
comp_tab<-cbind(overall,bmi_quan)
print(comp_tab,
      rtitle = "Summary Statistics For Complete Cases",
      cnames = c("Overall", "BMI Quartile 1", "BMI Quartile 2", "BMI Quartile 3","BMI Quartile 4"))
```
```{r}
#for target population
overall_t<-summary_table(target_pop,summarytab)
bmi_quan_t<-summary_table(target_pop,summarytab, by=c("bmi_quan"))
comp_tab_t<-cbind(overall_t,bmi_quan_t)
print(comp_tab_t,
      rtitle = "Summary Statistics For Complete Cases",
      cnames = c("Overall", "BMI Quartile 1", "BMI Quartile 2", "BMI Quartile 3","BMI Quartile 4"))
```


```{r}
##Just Outcome and POI
model_1<-lm(BPSysAve~BMI,data=comp_cases)
summary(model_1)
```

```{r}
#I added the other covariates in groups, in a sequential pattern
#Demographic
model_2<-lm(BPSysAve~BMI+Gender+Race1+Education+Age,data=comp_cases)
summary(model_2)
```

```{r}
#health variables:DirectChol+Diabetes+PhysActive+SmokeNow
model_3<-lm(BPSysAve~BMI+Age+Gender+Race1+Education+DirectChol+Diabetes+PhysActiveDays+SmokeNow+AlcoholYear,data=comp_cases)
summary(model_3)

plot_model(model_3,type="diag",show.data = TRUE)
```

```{r}
#interaction terms: I know that total cholesteral tends to increase with age and there might be an interaction there. Also, I think age and sex might have an interaction (need to see if literature agrees)
# Age*gender
# gender*cholesterol
# alcohol*diabetes: neither are significant so not included
# BMI*race
# Diabetes*cholesterol
comp_cases %>%
     ggplot(aes(x=BMI, 
               y=BPSysAve, 
               color=Race1))+
     geom_smooth(method="lm")+geom_point()

comp_cases %>%
     ggplot(aes(x=DirectChol, 
               y=BPSysAve, 
               color=Diabetes))+
     geom_smooth(method="lm")+geom_point()

comp_cases %>%
     ggplot(aes(x=DirectChol, 
               y=BPSysAve, 
               color=Gender))+
     geom_smooth(method="lm")+geom_point()

comp_cases %>%
     ggplot(aes(x=Age, 
               y=BPSysAve, 
               color=Gender))+
     geom_smooth(method="lm")+geom_point()

comp_cases %>%
     ggplot(aes(x=DirectChol, 
               y=BPSysAve, 
               color=Race1))+
     geom_smooth(method="lm")

comp_cases %>%
     ggplot(aes(x=Age, 
               y=BPSysAve, 
               color=Gender))+
     geom_smooth(method="lm")

cat_plot(model = model_main, pred = Education, modx = Race1, data = comp_cases)
cat_plot(model = model_main, pred = Race1, modx = Gender, data = comp_cases)
```

```{r}
model_4<-lm(BPSysAve ~ BMI + Age + Gender + Race1 + Education + 
    DirectChol + Diabetes + PhysActiveDays + AlcoholYear + SmokeNow + 
    Age * Gender + Gender * PhysActiveDays  + BMI * Race1 + DirectChol * 
    Diabetes +  DirectChol * Race1 + AlcoholYear * Diabetes, data = comp_cases)
summary(model_4)


model_5<-lm(BPSysAve ~ BMI + Age + Gender + Race1 + Education + 
    DirectChol + Diabetes + PhysActiveDays + AlcoholYear + SmokeNow, data = comp_cases)
summary(model_5)
#plot_model(model_4,type="diag",show.data = TRUE)

#if only consider interaction with significant covariates:
model_6 <- lm(BPSysAve ~ BMI + Age + Gender + Race1 + Education + 
    DirectChol + Diabetes + PhysActiveDays + AlcoholYear + SmokeNow + BMI*Race1 + Age*Gender , data = comp_cases)
summary(model_6)
#consider model_6 as the main model

```

plots for BMI, confounders and effect modifier:
```{r}
library(patchwork)
p1 = ggplot(comp_cases, aes(x = Gender)) + geom_bar(fill = "light blue") + labs(y = "numbers", title = "Gender") + theme(plot.title=element_text(hjust=0.5))

p2 = ggplot(comp_cases, aes(x = Race1)) + geom_bar(fill = "light blue") + labs(y = "numbers", title = "Race") + theme(plot.title=element_text(hjust=0.5))

p3 = ggplot(comp_cases, aes(x = Education)) + geom_bar(fill = "light blue") + labs(y = "numbers", title = "Education Level") + theme(plot.title=element_text(hjust=0.5))

p4 = ggplot(comp_cases, aes(x = PhysActiveDays)) + geom_bar(fill = "light blue") + labs(y = "numbers", title = "Physical Activity Days") + theme(plot.title=element_text(hjust=0.5))

p5 = ggplot(comp_cases, aes(x = SmokeNow)) + geom_bar(fill = "light blue") + labs(y = "numbers ", title = "Smoking or not") + theme(plot.title=element_text(hjust=0.5))

p6 = ggplot(comp_cases, aes(x = Age)) + geom_histogram(fill = "light blue", bins = 30) + labs(y = "numbers", title = "Age") + theme(plot.title=element_text(hjust=0.5))

p7 = ggplot(comp_cases, aes(x = AlcoholYear)) + geom_histogram(fill = "light blue", bins = 10) + labs(y = "numbers", title = "Drinking day per year") + theme(plot.title=element_text(hjust=0.5))

p8 = ggplot(comp_cases, aes(x = DirectChol)) + geom_histogram(fill = "light blue", bins = 10) + labs(y = "numbers", title = "Direct HDL") + theme(plot.title=element_text(hjust=0.5))

p9 = ggplot(comp_cases, aes(x = BMI)) + geom_histogram(fill = "light blue", bins = 10) + labs(y = "numbers", title = "BMI") + theme(plot.title=element_text(hjust=0.5))

(p1 + p2 + p3) / (p4 + p5 + p6) / (p7 + p8 + p9)
```

Model Diagnostics:
Check linearity assumption: 
through partial regression plots, linearity assumption holds
```{r}
car::avPlots(model_6)

# check linear relationship between SBP and BMI:
model_par_1 <- lm(BPSysAve ~ Age + Gender + Race1 + Education + 
    DirectChol + Diabetes + PhysActiveDays + AlcoholYear + SmokeNow +  Age*Gender , data = comp_cases)
model_par_2 <- lm(BMI ~ Age + Gender + Race1 + Education + 
    DirectChol + Diabetes + PhysActiveDays + AlcoholYear + SmokeNow +  Age*Gender , data = comp_cases)

ggplot(comp_cases, aes(x = model_par_2$residuals, y = model_par_1$residuals)) + geom_point() + geom_smooth(method = lm) + labs( x = "resid.2", y = "resid.1", title = "Partial regression plot between BMI and SBP")  +
  theme(plot.title = element_text(hjust = 0.5)) 
```

```{r}
library(quantreg)
m_raw = rq(comp_cases$BPSysAve ~ comp_cases$BMI, tau = seq(0,1,by = 0.25))
plot(m_raw)
```
check independence assumption:
no significant positive or negative trend for autocorrelation
```{r}
lag.plot(m6.res, lag = 1, diag.col = "black", main = "Lag plot", col = "dark red")
```
normality test: basically not violated
as studentized residuals go extreme, those points no longer fit the straight line b/c there
are quite a lot influencial points with high residual (see in influentialPlot below)
```{r}
#residual plot
comp_cases$m6.res = rstudent(model_6)

p_n1 = ggplot(comp_cases, aes(x = m6.res)) + geom_histogram(bins = 30) + labs(x = "externally studentized residuals", y = "frequency", title = "distribution of  residuals") + theme(plot.title=element_text(hjust=0.5)) 
#qq plot:
p_n2 = ggplot(comp_cases, aes(sample = m6.res)) + stat_qq_band() + stat_qq_line() + stat_qq_point() + labs(x = "theoretical quantiles", y = "sample quantiles", title = "QQ plot") + theme(plot.title=element_text(hjust=0.5))

(p_n1+p_n2)
```

check equal variance assumption:
although model 6 didn't pass the bp test, by residual plot we still find homoscadisticity 
```{r}
#perform Breusch-Pagan Test: heteroscasticity exists in the model
#library(lmtest)
#bptest(model_6)
#car::spreadLevelPlot(model_6) # suggest a power transformation with -2
p_eq_var = ggplot() + 
  geom_point(aes(y = m6.res, x = model_6$fitted.values)) + 
  labs( x = "Predicted Value", y = "Residuals", title = "Residuals vs. fitted values") + 
  theme(plot.title = element_text(hjust = 0.5)) 
p_eq_var
```
Or if we claim heteroscasticity and turn for robust standard error and correct t-value and p-value in the model, we find no much difference of explanation from summary()
```{r}
library(sandwich)
library(robustbase)
summary(model_6)
coeftest(model_6, vcov = vcovHC(model_6, type = "HC1"))
```

influence diagnostics:
```{r}
library(stats)
influence.measures(model_6)
car::influencePlot(model_6, main = "Regression influence plot") # obs 817, 287, 686, 687, 940 contribute most influence

#high leverage points with cutoff: leverage > 0.1
comp_cases$m6.h = hatvalues(model_6)
high_lev = m6.h[which(m6.h>0.1)]
as.data.frame(comp_cases[comp_cases$m6.h > 0.1,])

# obs 287 (ID 56114) has the greatest impact on parameter estimate of model_6
m6.D=cooks.distance(model_6)
which.max(m6.D)

# obs (ID 67624) has the most influence on MSE:
m6.covratio=covratio(model_6)
which.max(m6.covratio*(1-m6.h))

```
outlier detection:
```{r}
outlier_1 = comp_cases[abs(comp_cases$m6.res)>3,] # 20 obs with extremely high or low sbp
outlier_2 = comp_cases[abs(comp_cases$m6.res)>4,] # 7 obs with SBP > 190: extreme sbp values
outlier_1
outlier_2
```